A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Merging from the old master branch

Now we proceed with the definition of mode-correctness.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align*}
    \synvar(\cdot)                                  & = \emptyset  \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align*}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align*}
    \MC_{\mathit{as}}( \cdot ) ={} & \top \\
    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
    \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
  \end{align*}
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
Finally, for a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.

\begin{remark}
  As this is a condition for \emph{typing rules} and not for raw terms or typing derivations, this property cannot be formalised without the concept of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which are properties that can still be expressed for individual systems even without a general definition of bidirectional type systems.
  Now that we have the notion of signature available, we can articulate mode-correctness above.
\end{remark}

\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}

\begin{proof}

\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}

We assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
This requirement for decidable equality is necessary, as type comparisons are ubiquitous throughout type synthesis.
To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
For clarity in presentation, however, we will continue to use named variables.

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  For every context $\Gamma$ and raw term $\isTerm{t}$ in mode $d$, 
  \begin{enumerate}
    \item if $\isTerm{t}$ is synthesising, then it is decidable whether there is a type~$A$ and a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
      \]
    \item if $\isTerm{t}$ is checking, then every type $A$ decides whether there is a derivation of
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{d}$ (instead of $\isTerm{t}$ alone).
  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$ to check.
  The last case \Rule{Op} is the interesting part of this theorem.
  \begin{description}
    \item[\SynRule{Var}:] If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
      Otherwise, if $\isTerm{x}$ is not in $\Gamma$ but $\Gamma |- \isTerm{x} \syn A$ is derivable, then by inversion $\isTerm{x}$ must be in $\Gamma$, a contradiction.
    \item[\SynRule{Anno}:] For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
      \begin{enumerate}
        \item If $\Gamma |- \isTerm{t} \chk A$ is derivable, then $\Gamma |- \isTerm{t \annotate A} \syn A$ is derivable.
        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- \isTerm{t \annotate A} \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
      \end{enumerate}

    \item[\ChkRule{Sub}:] If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{enumerate}
        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
          \begin{enumerate}
            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t \chk A$.
            \item if $A \neq B$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
          \end{enumerate}
        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
      \end{enumerate}

%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
%      \[
%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
%      \]
%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
%      \[
%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
%      \]
    \item[\Rule{Op}:]
      Suppose that $t$ is of them $\tmOpts$ for some mode-correct operation $o$ in $\Omega$.
      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.

@@

      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
    \end{enumerate}
  \end{claim}
  \end{description}
\end{proof}

>>>>

Now we proceed with the definition of mode-correctness.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                                  & = \emptyset  \\
    \label{eq:synvar2}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \label{eq:synvar3}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align}
    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) ={} & \top \\
    \label{eq:MC2} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
    \label{eq:MC3} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
  \end{align}
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
Finally, for a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.

\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from the requirement that each variable in a context is unique.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts : A$ contains a substitution $\rho$ to concrete types from the local context $\Xi$ of an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$.
      To prove that any two derivations for the same $\tmOpts$ has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on $X = \synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ by induction on the list:
      \begin{itemize}
        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is checking, then it holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_i]A_{i}^{\dir{d_i}}\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
      \end{itemize}
  \end{itemize}
\end{proof}

Uniqueness above is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system, which can involve selecting synthesis arguments whose type variables include the variables of the conclusion.
For instance, for a derivation of $\Gamma |- t\;u \syn B$, one would apply the hypothesis to the sub-derivation $\Gamma |- t \syn A \bto B$.
On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.

\begin{remark}
  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
  With the concept of signature now at our disposal, we can express mode-correctness as defined formally above.
\end{remark}
\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}

For bidirectional type synthesis, we assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
This assumption is necessary, as type comparisons are ubiquitous throughout type synthesis. 
To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
For clarity in presentation, however, we will continue to use named variables.

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A;
      \]
    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$ (instead of $\isTerm{t}$ alone).
  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$ to check.
  The last case \Rule{Op} is the interesting part of this theorem.
  \begin{itemize}
    \item \SynRule{Var}: If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
      Otherwise, if $\isTerm{x}$ is not in $\Gamma$ but $\Gamma |- \isTerm{x} \syn A$ is derivable, then by inversion $\isTerm{x}$ must be in $\Gamma$, a contradiction.
    \item \SynRule{Anno}: For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
      \begin{enumerate}
        \item If $\Gamma |- \isTerm{t} \chk A$ is derivable, then $\Gamma |- \isTerm{t \annotate A} \syn A$ is derivable.
        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- \isTerm{t \annotate A} \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
      \end{enumerate}

    \item \ChkRule{Sub}: If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{enumerate}
        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
          \begin{enumerate}
            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t \chk A$.
            \item if $A \neq B$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
          \end{enumerate}
        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
      \end{enumerate}

%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
%      \[
%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
%      \]
%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
%      \[
%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
%      \]
    \item \Rule{Op}:
      Suppose that $t$ is of them $\tmOpts$ for some mode-correct operation $o$ in $\Omega$.
      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.

@@

      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
    \end{enumerate}
  \end{claim}
  \end{itemize}
\end{proof}

### 2023-07-08 tex/sec7-future.tex
[ tex ] Merging from the old master branch

>>>>

\todo[inline]{We leave the problem of presenting bidirectional typing for more general and advanced language features as future work (such as those related to polymorphic types~\citep{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018}), which requires advancement in language formalisation~(\cref{sec:language-formalisation}) that is orthogonal to our work.}

%\subsubsection{Theories of abstract syntax with variable binding}
%\label{sec:theory-of-syntax}
%
%The aforementioned frameworks except \citeauthor{Gheri2020}'s are at least inspired by \varcitet{Fiore1999}{'s} initial semantics for abstract syntax with variable binding using category theory.
%The main idea is that the set of (untyped) abstract syntax trees for a language consists of
%\begin{enumerate*}
%  \item a family of sets $\Term_{\Gamma}$ of well-scoped terms under a context~$\Gamma$ with
%  \item variable renaming for a function $\sigma\colon \Gamma \to \Delta$ between variables acting as a functorial map from $\Term_{\Gamma}$ to $\Term_{\Delta}$, i.e.\ a presheaf $\Term\colon \mathbb{F} \to \mathsf{Set}$, and
%  \item an initial algebra $[\mathsf{v}, \mathsf{op}]$ on~$\Term$ given by the variable rule as a map $\mathsf{v}$ from the presheaf~$V$ of variables (i.e.\ the embedding $V\colon \mathbb{F} \hookrightarrow \mathsf{Set}$) to $\Term$ and other constructs as $\mathsf{op}\colon \mathbb{\Sigma}\Term \to \Term$ where $\mathbb{F}$ is the category of contexts, the functor $\mathbb{\Sigma}\colon \mathsf{Set}^\mathbb{F} \to \mathsf{Set}^\mathbb{F}$ encodes the arities of constructs, and the initiality amounts to structural recursion, i.e.\ \emph{term traversal}.
%\end{enumerate*}
%To put it succinctly, it is the free $\mathbb{\Sigma}$-algebra over the presheaf~$V$ of variables.
%
%Fortunately, constructing the initial algebra of terms in type theory boils down to defining an inductive type with a few constructors that align with the variable rule and a rule schema for language constructs specified by a signature~\citep{Fiore2022}.

%Substitution is also modelled categorically, but it does not play a role in this paper.%
%\todo{Reveal a bit about this paper?}

%\begin{remark} \label{re:type-signature}
%Most of existing theories treat types independently of terms, thereby excluding them from signatures.
%To the best of our knowledge, the only exception to this approach is found in the work of~\citet{Arkor2020}, which incorporates signatures for both terms and types.
%Interestingly, this inclusion is also critical for type synthesis for comparing a concrete type $N \bto N$ with an abstract type $A \bto B$, where $A$ and $B$ are type variables in a typing rule.
%\end{remark}
